---
title: "Short Topic Summary: for + if/else"
author: "Oğuzhan Işılay"
date: "2026-01-12"
output:
  html_document: default
  pdf_document: default
---

# Short Topic Summary: for + if/else

## 1. For Loop

In R, a **for loop** allows us to repeat the same operation for multiple values.

Example structure:

```{r}
items <- c(1,2,3)

for (i in items) {
  # operation to be performed on each item
}
```


items → a vector or list containing multiple values  
i → represents one value from the list during each iteration

The loop follows this logic:

Take each element inside **items**, one by one, and execute the code inside the loop.

Example vector:

```{r}
leaf_lengths <- c(3.2, 5.1, 4.0)
```


The function **c()** means *combine*.  
It combines multiple values into a vector.

Here we created a vector named **leaf_lengths**.

Now we can print each value using a loop:

```{r}
for (u in leaf_lengths) {
  print(u)
}
```


The computer will:

First iteration → print first value  
Second iteration → print second value  
Third iteration → print third value  

The loop stops when the vector ends.

---

### Creating an Empty Vector

Sometimes we need a place to store results.

```{r}
results <- c()
```


This creates an **empty vector**.

Daily life analogy:

Imagine an empty basket where we will place our results.

Now we process the data:

```{r}
for (u in leaf_lengths) {
  results <- c(results, u * 2)
  print(results)
}
```


What happens here?

Each value of **leaf_lengths** is taken one by one.  
Each value is multiplied by 2.  
The result is added to the **results** vector.

So the loop is:

Take value → process it → store the result.

This structure is very common in biological data analysis.

---

# 2. if / else

The **if / else** structure allows the computer to make decisions.

Basic logic:

If something is true → do one thing  
Otherwise → do another thing

General structure:

'if (condition) {
  # code if true
}'

The condition must return **TRUE** or **FALSE**.

Example:

```{r}
leaf_length <- 6

if (leaf_length > 5) {
  print("Long leaf")
}

```

Since 6 > 5 is TRUE, the message is printed.

If the value was 4, the condition would be FALSE and nothing would happen.

---

### if / else structure

'if (condition) {
  # if true
} else {
  # if false
}'

Example:

```{r}
leaf_length <- 4

if (leaf_length > 5) {
  print("Long leaf")
} else {
  print("Short leaf")
}

```

Since 4 > 5 is FALSE, the **else** block runs.

Output:

Short leaf

---

### Multiple Conditions

Sometimes we need more than two conditions.

```{r}
if (leaf_length < 3) {
  print("Very short")
} else if (leaf_length < 6) {
  print("Medium")
} else {
  print("Long")
}

```

The computer checks conditions step by step.

First condition  
If FALSE → move to next condition  
Continue until one condition becomes TRUE.

---

# 3. Using for with if / else

Often we combine **loops** and **conditions**.

for → processes values one by one  
if / else → decides what to do with each value

Example:

```{r}
leaves <- c(2.5, 4.8, 6.1)

for (u in leaves) {
  if (u > 5) {
    print("Long leaf")
  } else {
    print("Short leaf")
  }
}

```

The logic is:

Take a leaf  
Check its length  
Print the classification  
Move to the next leaf

This represents the core logic behind many biological data analyses.

---

# Mini Example: GC% and Motif Analysis in DNA Sequences

## Example 1: Checking GC Content in a DNA Sequence

We examine whether each nucleotide forms a **GC strong bond** or an **AT weak bond**.

```{r}
DNA_sequence <- c("A", "G", "C", "T", "G")

for (nucleotide in DNA_sequence) {

  if (nucleotide == "G" | nucleotide == "C") {
    print(paste(nucleotide, ": GC Strong Bond"))
  } else {
    print(paste(nucleotide, ": AT Weak Bond"))
  }

}
```


---

## Example 2: Simple Motif Search

We have several DNA regions and want to detect the **start codon (ATG)**.

```{r}
regions <- c("CCG", "ATG", "TTA", "ATG")

for (i in 1:length(regions)) {

  if (regions[i] == "ATG") {
    print(paste(i, ". region contains a start codon!"))
  } else {
    print(paste(i, ". region is clean."))
  }

}
```


---

## Example 3: GC Ratio Calculation

We count how many **G** and **C** bases exist in the sequence.

```{r}
dna <- c("G", "C", "A", "T", "G", "G", "A")

gc_number <- 0

for (n in dna) {
  if (n == "G" | n == "C") {
    gc_number <- gc_number + 1
  }
}

total_length <- length(dna)

gc_ratio <- (gc_number / total_length) * 100

print(paste("Total GC Count:", gc_number))
print(paste("GC Ratio:", gc_ratio, "%"))
```


---

## Example 4: Multiple Conditions (Temperature and Enzyme Activity)

We evaluate enzyme activity at different temperatures.

```{r}
temperatures <- c(10, 37, 55, 90)

for (t in temperatures) {

  if (t < 20) {
    print(paste(t, "degrees: Enzyme is very slow."))
  } else if (t >= 20 & t <= 45) {
    print(paste(t, "degrees: Optimum working temperature."))
  } else {
    print(paste(t, "degrees: Enzyme structure may be damaged (Denaturation)."))
  }

}
```

